---
title: "Chapter 4 Data Structures"
format: html
editor: visual
---

## Data Structures

One of R’s most powerful features is its ability to deal with tabular data - such as you may already have in a spreadsheet or a CSV file. Let’s start by making a toy dataset in your `data/` directory, called `feline-data.csv`:

```{r}
cats <- data.frame(coat = c("calico", "black", "tabby"),
                    weight = c(2.1, 5.0, 3.2),
                    likes_string = c(1, 0, 1))
```

We can now save `cats` as a CSV file. It is good practice to call the argument names explicitly so the function knows what default values you are changing. Here we are setting `row.names = FALSE`. Recall you can use `?write.csv` to pull up the help file to check out the argument names and their default values.

```{r}
write.csv(x = cats, file = "data/feline-data.csv", row.names = FALSE)
```

The contents of the new file, `feline-data.csv`:

```{r}
coat,weight,likes_string
calico,2.1,1
black,5.0,0
tabby,3.2,1
```
